GSE158422: DNA methylation in Lung squamous cell carcinoma - RNA expression and integration

### 2023-09-27 figs/GSE158422_gmea.Rmd
more work on integrating methylation and RNA expression

@@

## GMEA

@@

## LA: limma aggregate 1-sample t-test

@@

head(dma) %>%
  kbl(caption="Top DM probes") %>%
  kable_styling("hover",full_width=FALSE)

@@

@@

# need to get the median column before analysis
m <- dmagg$gme_res_df[,"median",drop=FALSE]

tic()
lares <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
toc()

@@

## AL: Aggregate limma t-test

@@

mymed <- function(g) {
  probes <- rownames(myann[grep(g,myann$gnl),])
  rows <- which(rownames(mx) %in% probes)
  if ( length(rows) > 1 ) {
    b <- mx[rows,]
    med <- apply(b,2,median)
    med <- matrix(med,nrow=1)
    colnames(med) <- colnames(b)
    rownames(med) <- g
    return(med)
  }
}

@@

@@

#write.table(dma3a,file="dma3a.tsv")

@@

@@

@@

>>>>

boxplot(list("tumor"=matrix(colMeans(mval),ncol=2)[,1],"normal"=matrix(colMeans(mval),ncol=2)[,2]),
  main="mean probe methylation mval")

@@

# GMEA: all probes

@@

## LA: limma aggregate 1-sample t-test all probes

@@

hist(dm$t,breaks=30,main="all probes")
hist(subset(dm,Relation_to_Island=="Island")$t,main="CpG island")
hist(subset(dm,Relation_to_Island=="N_Shore")$t,main="CpG shore")
hist(subset(dm,Relation_to_Island=="N_Shelf")$t,main="CpG shelf")

hist(subset(dm,Regulatory_Feature_Group=="",main="No genomic feature")
hist(subset(dm,Regulatory_Feature_Group=="Promoter_Associated")$t,main="promoter")
hist(subset(dm,Regulatory_Feature_Group=="Gene_Associated")$t,main="gene")
hist(subset(dm,Regulatory_Feature_Group=="Unclassified")$t,main="unclassified")

head(dm) %>%
  kbl(caption="Top DM probes") %>%
  kable_styling("hover",full_width=FALSE)

@@

hist(dmagg$gme_res_df$mean,breaks=30) ; grid()

@@

# try mean and median
m <- dmagg$gme_res_df[,"mean",drop=FALSE]
lares <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
hist(lares$t_mean)

# need to get the median column before analysis
m <- dmagg$gme_res_df[,"median",drop=FALSE]
lares <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
hist(lares$t_median)

@@

## AL: Aggregate limma t-test all probes

@@

# compare median and mean based aggregation
mymed <- function(g) {
  probes <- rownames(myann[grep(g,myann$gnl),])
  rows <- which(rownames(mx) %in% probes)
  if ( length(rows) > 1 ) {
    b <- mx[rows,]
    med <- apply(b,2,median)
    med <- matrix(med,nrow=1)
    colnames(med) <- colnames(b)
    rownames(med) <- g
    return(med)
  }
}

@@

boxplot(list("normal"=matrix(colMeans(mval),ncol=2)[,2],"tumor"=matrix(colMeans(mval),ncol=2)[,1]),
  main="mean mval for control and tumor samples (all probes)")

boxplot(list("normal"=matrix(colMeans(medf),ncol=2)[,2],"tumor"=matrix(colMeans(medf),ncol=2)[,1]),
  main="mean mval for control and tumor samples (gene associated)")

@@

hist(magg$t)

@@

Very curious that the distribution is negatively skewed now.

Let's do another test to see whether this is correct or not.

```{r,test}

mytest <- lapply(1:nrow(medf),function(i) {
  ctl <- matrix(medf[i,],ncol=2)[,2]
  tum <- matrix(medf[i,],ncol=2)[,1]
  tt <- t.test(tum,ctl)
  t=tt$statistic ; p=tt$p.value ; est=rev(tt$estimate)
  names(est) = c("ctl","tum")
  c(t,p,est)
} )

mytest <- as.data.frame(do.call(rbind,mytest))
colnames(mytest) = c("t","p","ctl","tum")
rownames(mytest) <- rownames(medf)
hist(mytest[,"t"])

mytest <- mytest[order(mytest$p),]
head(mytest,20) %>%
  kbl(caption="Top significant differential gene methylation results using dumb test") %>%
  kable_paper("hover", full_width = F)

nrow(subset(mytest,p<0.05 & t>0))
nrow(subset(mytest,p<0.05 & t<0))

Hypomethylated genes still outnumber hypermethylated genes.

It does skew downwards after aggregation.

@@

hist(alres$t_mean)

@@

# GMEA: promoter probes only

## Limma-aggregate-1-sample t-test promoter only

```{r,ls_promoter}

dmp <- dm[grep("Promo",dm$Regulatory_Feature_Group),]

tic()
dmpagg <- agg(dmp,cores=CORES)
toc()

head(dmpagg$gme_res_df,20) %>%
  kbl(caption="top gmea genes (t-test)") %>%
  kable_paper("hover", full_width = F)

```{r,laphist}

hist(dmpagg$gme_res_df$median,breaks=30) ; grid()

```{r,t_enrichment_test1_promoter}

# need to get the median column before analysis
m <- dmpagg$gme_res_df[,"median",drop=FALSE]

tic()
lapres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
toc()

head(lapres,30) %>%
  kbl(caption = "Top significant genesets using LA approach for promoters") %>%
  kable_paper("hover", full_width = F)

head(lapres[order(-abs(lapres$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets using LA approach") %>%
  kable_paper("hover", full_width = F)

lapsig <- subset(lapres,fdr<0.05)
nrow(lapsig)

head(lapsig[order(-abs(lapsig$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets after FDR filtering using LA approach for promoters") %>%
  kable_paper("hover", full_width = F)

plot(lapres$t_median,-log10(lapres$pval),pch=19,col="gray")
grid()
points(lapsig$t_median,-log10(lapsig$pval),pch=19,col="red")

## AL: Aggregate limma t-test Promoter only

Determine the median of probe values.

```{r,med1p}

gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
gnl <- mclapply(gnl,unique,mc.cores=CORES)
myann$gnl <- gnl
myannp <- myann[grep("Promot",myann$Regulatory_Feature_Group),]

mxp <- mval[rownames(mval) %in% rownames(myannp),]

mymed <- function(g) {
  probes <- rownames(myannp[grep(g,myannp$gnl),])
  rows <- which(rownames(mxp) %in% probes)
  if ( length(rows) > 1 ) {
    b <- mxp[rows,]
    med <- apply(b,2,median)
    med <- matrix(med,nrow=1)
    colnames(med) <- colnames(b)
    rownames(med) <- g
    return(med)
  }
}

medp <- mclapply(gn,mymed,mc.cores=CORES)
medp <- medp[lapply(medp,length)>0]
medfp <- do.call(rbind,medp)

```{r,limma1p}

fit.reduced <- lmFit(medfp,design)
fit.reduced <- eBayes(fit.reduced)
summary(decideTests(fit.reduced))
maggp <- topTable(fit.reduced,coef=ncol(design), number = Inf)

head(maggp, 30) %>%
  kbl(caption="limma gene methylation results") %>%
  kable_paper("hover", full_width = F)

1-sample t-test for enrichment promoter.

```{r,ttestp}

m <- as.data.frame(maggp$t)
rownames(m) <- rownames(maggp)
colnames(m) <- "t"

alpres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)

alpsig <- subset(alpres,fdr<0.05)

nrow(alpsig)

nrow(subset(alpsig,t_median>0))

head(alpsig[order(-alpsig$t_median),],20) %>%
  kbl(caption="hypermethylated pathways") %>%
  kable_paper("hover", full_width = F)

nrow(subset(alpsig,t_median<0))

head(alpsig[order(alpsig$t_median),],20) %>%
  kbl(caption="hypomethylated pathways") %>%
  kable_paper("hover", full_width = F)

## Promoter Overlap

```{r,ol1p}

lapsig_up <- rownames(subset(lapsig,t_median>0))
lapsig_dn <- rownames(subset(lapsig,t_median<0))

alpsig_up <- rownames(subset(alpsig,t_median>0))
alpsig_dn <- rownames(subset(alpsig,t_median<0))

v1 <- list("LA up"=lapsig_up, "LA dn"=lapsig_dn,
  "AL up"=alpsig_up,"AL dn"=alpsig_dn)

### 2023-09-27 figs/GSE158422_rna_int.Rmd
more work on integrating methylation and RNA expression

human=x$row.names
checkGeneSymbols(human)
fix <- checkGeneSymbols(human)
rownames(x) <- fix$Suggested.Symbol
x$gene <- fix$Suggested.Symbol
x <- aggregate(. ~ gene,x,sum)
dim(x)
row.names(x)=x$gene
x$gene=NULL
head(x)

@@

sig <- subset(dge,padj<0.05)
SIG = nrow(sig)
DN = nrow(subset(sig,log2FoldChange<0))
UP = nrow(subset(sig,log2FoldChange>0))
HEADER = paste("mRNA", SIG , "DEGs,", UP ,"upregulated,", DN, "downregulated")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3,
  xlab="log2 base mean",
  ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 ,xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red"
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- rep(c(0,1),ncol(x)/2)
colCols <- gsub("0","gray",colCols)
colCols <- gsub("1","black",colCols)

@@

head(lsres)

@@

>>>>

human=x$row.names
fix <- checkGeneSymbols(human)
x$gene <- fix$Suggested.Symbol
x$row.names=NULL
x <- aggregate(. ~ gene,x,sum)
dim(x)
row.names(x)=x$gene
x$gene=NULL
head(x)

@@

sig <- subset(dge,padj<0.05)
SIG = nrow(sig)
DN = nrow(subset(sig,log2FoldChange<0))
UP = nrow(subset(sig,log2FoldChange>0))
HEADER = paste("mRNA", SIG , "DEGs,", UP ,"upregulated,", DN, "downregulated")
# smear
plot(log2(dge$baseMean),dge$log2FoldChange,cex=0.6,cex.axis=1.2,cex.lab=1.3,
  xlab="log2 base mean",
  ylab="log2 fold change" ,pch=19,col="#838383")
points(log2(sig$baseMean),sig$log2FoldChange,cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
top <- head(sig,20)
# volcano
plot(dge$log2FoldChange, -log2(dge$pvalue) ,cex=0.6, cex.lab=1.3,cex.axis=1.2,
 ,xlab="log2 fold change", ylab="-log2 p-value" ,pch=19,col="#838383")
points(sig$log2FoldChange, -log2(sig$pvalue),cex=0.6,pch=19,col="red")
mtext((HEADER),cex=1.2)
# top N gene heatmap
colCols <- rep(c(0,1),ncol(x)/2)
colCols <- gsub("0","gray",colCols)
colCols <- gsub("1","black",colCols)

@@

head(lares)

@@

## Load promoter methylation enrichment data and calculate overlap

```{r,overlap2}

head(alpres)

head(lapres)

lapsig_up <- rownames(subset(lapsig,t_median>0))
lapsig_dn <- rownames(subset(lapsig,t_median<0))

alpsig_up <- rownames(subset(alpsig,t_median>0))
alpsig_dn <- rownames(subset(alpsig,t_median<0))

v1 <- list("LAP up"=lapsig_up, "LAP dn"=lapsig_dn,
  "RNA up"=msig_up,"RNA dn"=msig_dn)
plot(euler(v1),quantities = TRUE)

v1 <- list("ALP up"=alpsig_up,"ALP dn"=alpsig_dn,
  "RNA up"=msig_up,"RNA dn"=msig_dn)
plot(euler(v1),quantities = TRUE)

## Conclusion

Although there were some differentially methylated pathways analysed with different methods,
they were not concordant with differential expression.
